---
title: Supplementary Materials for Literature mining the roles of genes in cancer to enable personalized cancer treatment
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::word_document2
bibliography: bibliography.bibtex
csl: nature-methods.csl
---

```{r cancermineSetup, include=FALSE}
source('cancermine/dependencies.R')

knitr::opts_chunk$set(echo = TRUE)
```

```{r cancermineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('cancermine/fig_annotationAndModel.R')
source('cancermine/fig_textminedOverview.R')
source('cancermine/fig_time.R')
source('cancermine/fig_accrualAndDualRoles.R')
source('cancermine/fig_comparisons.R')
source('cancermine/fig_comparisons_lowthreshold.R')
source('cancermine/fig_clustersAndTCGA.R')
invisible(dev.off())
```

```{r suptab_filterterms, echo=F, eval=T}
filtertermsTable <- data.frame(Term=c('tumor suppress','oncogen','driv'))

suppressWarnings(filtertermsTable %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Filter terms to identify sentences that are more likely to discuss the role of a gene in a cancer.',
    col.names = c('Filter Term')
    ) %>%
  kable_styling())

```




```{r suptab_interannotator, echo=F, eval=T}
interannotatorAgreement <- read.table('cancermine/annotation.interannotator.tsv', sep='\t', header=T, row.names=1)
interannotatorAgreement <- round(interannotatorAgreement,3)
rownames(interannotatorAgreement) <- c('Annotator 1','Annotator 2','Annotator 3')
colnames(interannotatorAgreement) <- c('Annotator 1','Annotator 2','Annotator 3')

suppressWarnings(interannotatorAgreement %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Annotation agreement between the three annotators measured by F1-score.'
    ) %>%
  kable_styling())

```

```{r suptab_examples, echo=F, eval=T}
exampleSentences <- read.table('cancermine/examples.txt', sep='\t', header=T)

suppressWarnings(exampleSentences %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Examples of sentences that contain mentions of drivers, oncogenes or tumor suppressors and should be extracted by the CancerMine system.',
    col.names = c('Association Type','PubMed ID','Sentence')
    ) %>%
  kable_styling())

```


```{r suptab_performance, echo=F, eval=T}
perfTableRounded <- perfTable
perfTableRounded$precision <- round(perfTableRounded$precision,3)
perfTableRounded$recall <- round(perfTableRounded$recall,3)
suppressWarnings(perfTableRounded %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'The thresholds and associated performance statistics for the three cancer-gene association types based on cross-validation.',
    col.names = c('Association Type','Threshold','Precision','Recall')
    ) %>%
  kable_styling())

```

```{r suptab_erroranalysis, echo=F, eval=T}
erroranalysis <- read.table('cancermine/error_analysis.tsv',header=T,sep='\t', stringsAsFactors=F)

suppressWarnings(erroranalysis %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Error analysis of the relations in the test set. This shows a confusion matrix of actual test data (down left) and their predictions (across top). There is no cross-classification where a relation is mislabelled as another type of association. The only errors are from the expected false negatives.',
    col.names = c('','Driver','Oncogene','Tumor Suppressor','negative')
    ) %>%
  kable_styling())
```

```{r suptab_novelexamples, echo=F, eval=T}
exampleSentences <- read.table('cancermine/novelExamples.txt', sep='\t', header=T)

suppressWarnings(exampleSentences %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Examples of sentences that are found to be novel and not novel that are extracted from different sections of the paper. Frequently, the mentions in Introductions discuss previous work, while the mentions in the Results reflect on the results of this paper and not previous papers. However, sometimes mentions are not found and are not correctly identified as novel.',
    col.names = c('Is Novel?','Section in Paper','Role','Gene','Cancer','PubMed ID','Sentence')
    ) %>%
  kable_styling())

```


```{r suptab_dualroles, echo=F, eval=T}
dualRoleTable <- as.data.frame(genesThatAreBothWithCancerTypes)
#dualRoleTable$gene <- as.character(dualRoleTable$gene)
#dualRoleTable$oncogenic <- factor(genesThatAreBothWithCancerTypes$oncogenic)
#dualRoleTable$tumorSuppressive <- factor(genesThatAreBothWithCancerTypes$tumorSuppressive)
#dualRoleTable <- dualRoleTable[,c('gene','oncogenic')]
dualRoleTable$oncogenic <- gsub("\n",", ",dualRoleTable$oncogenic)
dualRoleTable$tumorSuppressive <- gsub("\n",", ",dualRoleTable$tumorSuppressive)

suppressWarnings(dualRoleTable %>%
  #dplyr::mutate_all(linebreak) %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Some genes can act as an oncogene in one cancer and a tumor suppressor in another cancer. CancerMine can find these genes and this table shows a selection that have strong support (at least 4 papers and >90% certainty of their role in a cancer type).',
    col.names = c('Gene','Oncogene In','Tumor Suppressor In'),
    row.names = F
    ) %>%
  kable_styling())

```

```{r supfig_subsectionnovelty, echo=F, eval=T, fig.width=15, fig.asp=0.5,  fig.cap='(ref:supfig_subsectionnovelty)'}
grid.arrange(subsectionPlotWithNovelty)
```

(ref:supfig_subsectionnovelty) The link between a cancer gene associations novelty to CancerMine and the location within the paper where it is founded is shown. Mentions in the Introduction are much more likely to be not novel compared to other sections.


```{r supfig_sections, echo=F, eval=T, fig.width=15, fig.asp=0.5,  fig.cap='(ref:supfig_sections)'}
grid.arrange(sectionPlotWithoutHeader)
```

(ref:supfig_sections) The mentions of drivers, oncogenes and tumor suppressors are increasingly being extracted from the main section of full-text articles. Please note that the minor dip in 2018 numbers (compared to 2017) is due to many 2018 papers only becoming accessible for text-mining later into 2019.


```{r supfig_comparisons_lowthreshold, echo=F, eval=T, fig.width=11, fig.asp=0.5,  fig.cap='(ref:supfig_comparisons_lowthreshold)'}
grid.arrange(fig_comparisons_lowthreshold)
```

(ref:supfig_comparisons_lowthreshold) To explore the effect of high-precision low-recall classifier, we show a comparison without the strict thresholding. All classifiers use 0.5 instead of the higher thresholds. There is still not substantial overlap with the CGC and IntOGen resources suggesting that many of the gene associations in these databases are not mentioned in the literature in a form that can be extracted by CancerMine.


```{r supfig_tcga, echo=F, eval=T, fig.width=15, fig.asp=0.5,  fig.cap='(ref:supfig_tcga)'}
grid.arrange(fig_tcgaWithoutHeader)
```

(ref:supfig_tcga) All samples in seven TCGA projects are analyzed for likely loss-of-function mutations compared with the CancerMine tumor suppressor profiles and matched with the closest profile. Percentages shown in each cell are the proportion of samples labeled with each CancerMine profile that are from the different TCGA projects. Samples that match no tumor suppressor in these profiles or are ambiguous are assigned to none. The TCGA projects are breast cancer (BRCA), colorectal adenocarcinoma (COAD), liver hepatocellular carcinoma (LIHC), prostate adenocarcinoma (PRAD), low grade glioma (LGG), lung adenocarcinoma (LUAD) and stomach adenocarcinoma (STAD).

